---
layout: post
title: "Popularity of the Names of Disney Princesses"
date: 2016-04-12 14:00:00
jumbo_title: Popularity of the Names of Disney Princesses
jumbo_subtitle:
---

This is a small exploration of the `babynames` package which
contains a dataset tracking how many children were born with
a certain name according to the US Social Security 
Administration. I thought it would be interesting to look at
the popularity of the names of the Disney Princesses. Spikes
in popularity roughly correspond to when their respective 
movies were released.

```{r, message=FALSE, fig.width=10, fig.asp=0.5625}
library(dplyr)
library(babynames)

# I can show you the world...

jasmine <- babynames %>%
  filter(name == "Jasmine", sex == "F")

ariel <- babynames %>%
  filter(name == "Ariel", sex == "F")

belle <- babynames %>%
  filter(name == "Belle", sex == "F")

tiana <- babynames %>%
  filter(name == "Tiana", sex == "F")

merida <- babynames %>%
  filter(name == "Merida", sex == "F")

elsa <- babynames %>%
  filter(name == "Elsa", sex == "F")
```

```{r, message=FALSE, fig.width=10, fig.asp=0.5625}
plot(jasmine$year, jasmine$prop, type = "l", lwd = 2,
     col = "Purple", xlim = c(1880, 2015), xlab = "Year", 
     ylab = "Proportion of Baby Names", bty = "n")

lines(ariel$year, ariel$prop, type = "l", lwd = 2,
     xlab = "Year", ylab = "Proportion of Baby Names")

lines(belle$year, belle$prop, type = "l", lwd = 2,
     col = "#FFCC33")

lines(tiana$year, tiana$prop, type = "l", lwd = 2,
     col = "green")

lines(merida$year, merida$prop, type = "l", lwd = 2,
     col = "red")

lines(elsa$year, elsa$prop, type = "l", lwd = 2,
     col = "blue")

legend("topleft", legend = c("Jasmine", "Ariel", "Belle",
                             "Tiana", "Merida", "Elsa"),
       lwd = rep(2, 6), col = c("Purple", "Black", "#FFCC33",
                                "Green", "Red", "Blue"),
       bty = "n")
```

```{r, message=FALSE, fig.width=10, fig.asp=0.5625}
# Shining, shimmering, splendid

plot(belle$year, belle$prop, type = "l", lwd = 3,
      col = "#FFCC33", xlim = c(1970, 2015),
     ylim = c(0, max(elsa$prop)), bty = "n",
     xlab = "Year", 
     ylab = "Proportion of Baby Names")

lines(tiana$year, tiana$prop, type = "l", lwd = 3,
      col = "green")

lines(merida$year, merida$prop, type = "l", lwd = 3,
      col = "red")

lines(elsa$year, elsa$prop, type = "l", lwd = 3,
      col = "blue")

legend("topleft", legend = c("Belle",
                             "Tiana", "Merida", "Elsa"),
       lwd = rep(2, 4), col = c("#FFCC33",
                                "Green", "Red", "Blue"),
       bty = "n")

# Tell me data scientist, now when did you last let your model decide?
```