---
layout: post
title: "Drone Visualization Challenge"
date: 2016-04-05 20:00:00
jumbo_title: Drone Visualization Challenge
jumbo_subtitle:
---

I'm throwing my hat into the ring with this very last
minute submission to [Bob Rudis](https://twitter.com/hrbrmstr)'
weekly visualization [challenge](https://rud.is/b/2016/03/30/introducing-a-weekly-r-python-js-etc-vis-challenge/). First a bit of self-promotion using the minimap
package:

```{r, message=FALSE, fig.width=10, fig.asp=0.5625}
library(dplyr)
library(readxl)
library(minimap)
library(magrittr)
library(RColorBrewer)
library(stringr)
library(lubridate)

drone1 <- read_excel("~/Developer/R/drones/UASEventsNov2014-Aug2015.xls")
drone2 <- read_excel("~/Developer/R/drones/UAS_Sightings_report_21Aug-31Jan.xlsx")

drone <- data.frame(DT = c(drone1$EventDATETIME, drone2$EventDATETIME),
                    City = c(drone1$LocationCITY, drone2$LocationCITY),
                    State = c(drone1$LocationSTATE, drone2$LocationSTATE),
                    stringsAsFactors = FALSE)

state_name <- c(state.name, "District of Columbia")

drone %<>%
  arrange(desc(DT)) %>%
  filter(State %in% state_name)

drone$State_abb <- sapply(drone$State, function(x){
  usa_abb[grep(paste0("^", x, "$"), state_name)]
})

drone_by_state <- drone %>%
  group_by(State_abb) %>%
  summarize(n = n()) %>%
  mutate(col = colorRampPalette(c("#FCFBFD", "#3F007D"))(299)[n])

drone_by_state <- rbind(drone_by_state, data.frame(State_abb = setdiff(usa_abb, drone_by_state$State_abb), n = rep(0, 4), col = rep("#FFFFFF", 4), stringsAsFactors = FALSE))
drone_by_state$name_col <- sapply(drone_by_state$col, function(x){
  color <- col2rgb(x)
  if((color[1]*0.299 + color[2]*0.587 + color[3]*0.114) > 186){
    "#000000"
  } else {
    "#ffffff"
  }
})

opar <- par(no.readonly = TRUE)
par(xpd = TRUE)
miniusa(drone_by_state$State_abb, drone_by_state$col, 
        state_name_colors = drone_by_state$name_col)
mtext("Drone Sightings by State", cex = 1.5, line = -1)
legend(10, 2, c("50", "100", "200", "300"), pch = rep(15, 4),
       col = colorRampPalette(c("#FCFBFD", "#3F007D"))(299)[c(50, 100, 200, 299)], bty = "n")
```

Then I wondered: what kinds of weather conditions do drone
operators like to fly in?

```{r, eval=FALSE}
library(rnoaa)

# Get your token here: http://www.ncdc.noaa.gov/cdo-web/token
options(noaakey = "")

locations1 <- ncdc_locs(locationcategoryid='CITY', 
                       sortfield='name', sortorder='desc', 
                       limit = 1000)
locations2 <- ncdc_locs(locationcategoryid='CITY', 
                        sortfield='name', sortorder='desc', 
                        limit = 1000, offset = 1001)
locations <- rbind(locations1$data, locations2$data, 
                   stringsAsFactors = FALSE)

locations %<>%
  mutate(City = regmatches(name, regexpr("[^,]*", name))) %>%
  filter(grepl("CITY:US", id)) %>%
  mutate(State_abb = regmatches(name, regexpr("[^,]*$", name))) %>%
  mutate(State_abb = substr(State_abb, 2, 3)) %>%
  select(id, City, State_abb)

drone %<>%
  left_join(locations)
  
drone_weather <- drone %>%
  filter(!is.na(id)) %>%
  mutate(date = as.character(as.Date(DT)))
drone_weather$tmax <- rep(NA, nrow(drone_weather))
drone_weather$tmin <- rep(NA, nrow(drone_weather))

response <- list()

for(i in 1:nrow(drone_weather)){
  response[[i]] <- ncdc(datasetid='GHCND', locationid = drone_weather$id[i],
                   startdate = drone_weather$date[i], datatypeid = c("TMAX", "TMIN"),
                   enddate = drone_weather$date[i], limit=500)
}

for(i in 1:length(response)){
  drone_weather$tmax[i] <- mean(response[[i]]$data$value[response[[i]]$data$datatype == "TMAX"])
  drone_weather$tmin[i] <- mean(response[[i]]$data$value[response[[i]]$data$datatype == "TMIN"])
}
```

```{r, echo=FALSE, message=FALSE}
locations <- readRDS("~/Developer/R/drones/locations.rds")
response <- readRDS("~/Developer/R/drones/bob.rds")

drone %<>%
  left_join(locations)
  
drone_weather <- drone %>%
  filter(!is.na(id)) %>%
  mutate(date = as.character(as.Date(DT)))
drone_weather$tmax <- rep(NA, nrow(drone_weather))
drone_weather$tmin <- rep(NA, nrow(drone_weather))

for(i in 1:length(response)){
  drone_weather$tmax[i] <- mean(response[[i]]$data$value[response[[i]]$data$datatype == "TMAX"])
  drone_weather$tmin[i] <- mean(response[[i]]$data$value[response[[i]]$data$datatype == "TMIN"])
}
```

Due to data being unavailable and personal lack of know-how
there are only `r length(response)` drone sightings which
have temperate data.

```{r, fig.keep="none"}
drone_weather %<>%
  filter(!is.na(tmin)) %>%
  mutate(tmax = tmax/10, tmin = tmin/10)

h1 <- hist(drone_weather$tmax, breaks = 15)
h2 <- hist(drone_weather$tmin, breaks = 15)
```

```{r, fig.width=10, fig.asp=0.5625}
plot(h1, col = rgb(1, 0, 0, .6), main = "Temperature During Drone Sightings",
     xlab = "Temperature (C)", xlim = c(-5, 40))
plot(h2, col = rgb(0, 0, 1, .6), main = "Temperature During Drone Sightings",
     xlab = "Temperature (C)", add = TRUE)
legend(32, 20, c("Daily High", "Daily Low"), pch = rep(15, 2),
       col = c(rgb(1, 0, 0, .6), rgb(0, 0, 1, .6)), bty = "n")
```

```{r, fig.keep="none"}
drone_weather %<>%
  mutate(tmax = tmax*(9/5)+32, tmin = tmin*(9/5)+32)

h1 <- hist(drone_weather$tmax, breaks = 15)
h2 <- hist(drone_weather$tmin, breaks = 15)
```

```{r, fig.width=10, fig.asp=0.5625}
plot(h1, col = rgb(1, 0, 0, .6), main = "Temperature During Drone Sightings",
     xlab = "Temperature (F)", xlim = c(20, 110), ylim = c(0, 40))
plot(h2, col = rgb(0, 0, 1, .6), main = "Temperature During Drone Sightings",
     xlab = "Temperature (F)", add = TRUE)
legend(20, 20, c("Daily High", "Daily Low"), pch = rep(15, 2),
       col = c(rgb(1, 0, 0, .6), rgb(0, 0, 1, .6)), bty = "n")
```
